Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1) \note{\scriptsize Authors, please add/check middle initials and affiliations}

### 2014-04-13 WSSSPE1_report/WSSSPE1_summary_report.tex

\thanks{{}$^{(1)}$ National Science Foundation, Arlington, VA, USA; Computation Institute, University of Chicago \& Argonne National Laboratory, Chicago, IL, USA; \url{d.katz@ieee.org}}
%
\thanks{{}$^{(2)}$ National Evolutionary Synthesis Center (NESCent),
  Durham, NC, USA; \url{hlapp@nescent.org}}
%
\thanks{{}$^{(3)}$ Argonne National Laboratory, Chicago, IL, USA; \url{ketan@anl.gov}}
%
\thanks{{}$^{(4)}$ Center for Computation and Technology, Louisiana State University, Baton Rouge, LA, USA; \url{knarf@cct.lsu.edu}}
%
\thanks{{}$^{(5)}$ Columbia Astrophysics Laboratory, Columbia University, New
York, NY, USA; \url{matthewturk@gmail.com}}
%
\thanks{{}$^{(7)}$ San Diego Supercomputer Center, University of California San Diego, La Jolla, CA, USA, \url{wilkinsn@sdsc.edu}}
%
\thanks{{}$^{(8)}$ Leader, Research Software Development Team, University College London, London, UK; \url {j.hetherington@ucl.ac.uk}}
%
\thanks{{}$^{(9)}$ University of Texas at Austin, Austin, TX, USA; \url{jhowison@ischool.utexas.edu}}
%
\thanks{{}$^{(10)}$ NORC at the University of Chicago and the Illinois Institute of Technology, Chicago, IL, USA; \url{sctchoi@uchicago.edu}}
% 
\thanks{{}$^{(11)}$ xxx}
%
\thanks{{}$^{(12)}$ xxx}
%
\thanks{{}$^{(13)}$ xxx}
%
\thanks{{}$^{(14)}$ xxx}

\begin{abstract} %\note{Lead: Sou-Cheng T. Choi}
This document summarizes the First Workshop on Sustainable Software for Science:
Practice and Experiences (WSSSPE1). The content is based on the workshop
itself, as recorded in a set of collaborative notes taken during the workshop,
including discussion about the keynote presentations, three themed panels that
discussed papers accepted at the workshop`, other discussion, cross-cutting
issues, use cases, and conclusions. It joins a nascent literature seeking to
understand what drives software work in science and how the reward systems of
science work, thereby shaping the type of software work undertaken, including the
extent to which developers are motivated to build software for the long-term,
for the use of others, and whether to work collaboratively or separately.
Unique perspectives were captured about issues such as comprehensive documentation,
software deployment, and development practices; software licenses; career paths
for developers; identifiers for software through a
mechanism such as a Digital Object Identifier or publication of a ``software
paper''; and evidence of software contribution and impact, recorded in online
systems such as source code repositories like GitHub. For truly sustainable
software, there should be no endpoint, as the software products continue to be
used and useful beyond a single institution, grant, and developer or
development team.

@@

Because progress in scientific research is dependent on the quality of and
accessibility to software at all levels, it is now critical to address many
challenges related to the development, deployment, and maintenance of reusable
software.
%\note{Ketan: What are the levels in scientific process where software
%is required. It might be interesting to outline these stages and types of
%software required in each stage.}
In addition, it is essential that scientists,
researchers, and students are able to learn and adopt software-related skills
and methodologies. Established researchers are already acquiring some of these
skills, and in particular a specialized class of software developers is
emerging in academic environments who are an integral and embedded part of
successful research teams. The WSSSPE workshop  provided a forum
for discussion of the challenges around scientific software, including both positions and experiences. The
workshop short papers and debates have been archived to provide a basis for continued
discussion, and have fed into the collaborative writing of this
document. Additional papers based on extended versions of workshop
submissions have been developed as part of this special journal edition. The level of interest in the workshop which involved 
an estimated 90 to 150 participants during the day has led to further activities:
Two additional WSSSPE workshops are planned  (at the 2014 SciPy and SC14 conferences), 
funds to support the workshops have been obtained from 
the National Science Foundation and the Gordon and Betty Moore
Foundation, and the
original workshop website has been turned into a community website for
further discussion and progress. Additionally, a minisymposium  to
further explore some of the key issues raised in WSSSPE is co-organized by a
WSSSPE1 participant at the 2014 Society for Industrial and Applied Mathematics
(SIAM) Annual Meeting on ``Reliable
Computational Science.''

This paper summarizes the  workshop papers
and discussion, and identifies issues that future WSSSPE events will address.
Before the WSSSPE1 workshop took place, the organizers self-published
a paper~\cite{WSSSPE1-pre-report} to document the process of
organizing and advertising the workshop, collecting and reviewing the
papers, and putting together the agenda. Section~\ref{sec:process} is
based on that publication. The remainder of this paper is based on the
workshop itself, as documented in a set of collaborative notes taken
during the workshop~\cite{WSSSPE1-google-notes}, including discussion
about the keynote presentations (\S\ref{sec:keynotes}), the three
panels (\S\ref{sec:devel}-\ref{sec:community}), 
cross-cutting issues (\S\ref{sec:cross-cutting}),
use cases (\S\ref{sec:use-cases}), and conclusions
(\S\ref{sec:conclusions}).  Additionally, the accepted workshop papers are listed in
Appendix~\ref{sec:papers}, and a partial list of workshop attendees is
in Appendix~\ref{sec:attendees}.

@@

Another problem he experienced was staff retention with respect to
private organizations which reward those combining research and
software expertise (the ``Google Bus''). However, he is a strong fan
of software sustainability through public-private partnerships. He
noted that making a successful business from scientific software alone
is very rare: founders overvalue while customers undervalue. He noted
that to last, an open source project needs a minimal funding
requirement even with a vibrant community --- goodwill only goes so
far if one is being paid to do something else.  He talked about grant
schemes of relevance in the U.S., particularly with regard to
technology transfer~\cite{sbir-web, fased-web}.

He also had problems with selling research software: the university technology
transfer office wanted huge and unrealistic intellectual property
reach through, whereby they would get a share of profits from drugs
developed by pharmaceutical companies who use the software. 
He advocated for  a one-click approach for customers to
purchase university-written software.

@@

He concluded by discussing his 2011 paper ``Ten Simple Rules for Getting Ahead as a Computational Biologist in Academia''~\cite{bourne_ten} and
argued that computational scientists ``have a responsibility to
convince their institutions, reviewers, and communities that software is
scholarship, frequently
more valuable than a research article''.

@@

The question of what ``sustainable'' means in the context of software was
raised (see \S\ref{sec:defining-sustainability}.) The resources at OSS Watch~\cite{osswatch}   discuss what
to look for when choosing software for procurement, or reuse in further software
development. Regardless of the license and development model that will be used,
the future of the project must be considered. Even if a particular piece of
software satisfies today's requirements, will it continue to do so in the
future? Important questions include whether the supplier will still be around in five
years time, will it still care for its customers needs, will it be responsive to
bug reports and feature requests? Should you tie your investment to a single
supplier using a proprietary product, or ensure the project uses an OSI-approved
license, and can outlive any single entity if the software is still important.

@@

The papers submitted to this panel included many general
recommendations for processes, practices, tools, etc. One of the papers
suggested that a ``Software Sustainability Institute'' should be vested with
the authority to develop standardized quality processes, a central common
repository, central resources for services and consulting, a think tank of
sorts, and a software orphanage center. The idea of one common repository received
some resistance, with so many compelling alternatives available, e.g. Bitbucket or
GitHub. The point for centralization of communication/point of contact was seen as
reasonable, with the statement that ``vested with authority'' is perhaps too strong,
but ``providing tools if needed'' might be more appropriate.

@@

There was debate that while some of the basics of computing may be fairly
similar tools commonly used for computing have changed quite
significantly. Reference was made to Perl, which was commonly used, giving way
to whole new languages, such as Python, for gluing things together and how this
induces many students into entirely rewriting the scaffolding, leaving the old
to rot and the experiments to become non-reproducible as the tools change.
 There was
discussion of this tendency along with the enormous differences in the speed and
ease of sharing---having to ship tapes around (GCC, Emacs, etc) as opposed to
the immediate sharing of the latest development online, using revision control
systems like CVS, Subversion, Git, Mercurial, Bazaar, etc.

>>>>

\thanks{{}$^{(1)}$ National Science Foundation, Arlington, VA, USA; Computation Institute, University of Chicago \& Argonne National Laboratory, Chicago, IL, USA; \url{d.katz@ieee.org}}
%
\thanks{{}$^{(2)}$ National Evolutionary Synthesis Center (NESCent),
  Durham, NC, USA; \url{hlapp@nescent.org}}
%
\thanks{{}$^{(3)}$ Argonne National Laboratory, Chicago, IL, USA; \url{ketan@anl.gov}}
%
\thanks{{}$^{(4)}$ Center for Computation and Technology, Louisiana State University, Baton Rouge, LA, USA; \url{knarf@cct.lsu.edu}}
%
\thanks{{}$^{(5)}$ Columbia Astrophysics Laboratory, Columbia University, New
York, NY, USA; \url{matthewturk@gmail.com}}
%
\thanks{{}$^{(6)}$ Scientific Computing Group, Kitware, Inc., 28 Corporate Drive, Clifton Park, NY, USA; \url{marcus.hanwell@kitware.com}}
%
\thanks{{}$^{(7)}$ San Diego Supercomputer Center, University of California San Diego, La Jolla, CA, USA, \url{wilkinsn@sdsc.edu}}
%
\thanks{{}$^{(8)}$ Leader, Research Software Development Team, University College London, London, UK; \url {j.hetherington@ucl.ac.uk}}
%
\thanks{{}$^{(9)}$ University of Texas at Austin, Austin, TX, USA; \url{jhowison@ischool.utexas.edu}}
%
\thanks{{}$^{(10)}$ NORC at the University of Chicago and the Illinois Institute of Technology, Chicago, IL, USA; \url{sctchoi@uchicago.edu}}
% 
\thanks{{}$^{(11)}$ xxx}
%
\thanks{{}$^{(12)}$ xxx}
%
\thanks{{}$^{(13)}$ xxx}
%
\thanks{{}$^{(14)}$ xxx}

\begin{abstract} %\note{Lead: Sou-Cheng T. Choi}
This document summarizes the First Workshop on Sustainable Software for Science:
Practice and Experiences (WSSSPE1). The content is based on 
a set of collaborative notes taken during the workshop,
including discussion about the keynote presentations, three themed panels that
discussed papers accepted at the workshop`, other discussion, cross-cutting
issues, use cases, and conclusions. It joins a nascent literature seeking to
understand what drives software work in science and how the reward systems of
science function, thereby shaping the types of software projects undertaken, including the
extent to which developers are motivated to build software for the long-term,
for the use of others, and whether to work collaboratively or separately.
Unique perspectives were captured about issues such as comprehensive documentation, 
development and deployment practices; software licenses; career paths
for developers; identifiers for software through a
mechanism such as a Digital Object Identifier or publication of a ``software
paper''; and evidence of software contribution and impact, recorded in online
systems such as source code repositories like GitHub. For truly sustainable
software, there should be no endpoint, as the software products continue to be
used and useful beyond a single institution, grant, and developer or
development team.

@@

Because progress in scientific research is dependent on the quality of and
accessibility to software at all levels, it is now critical to address many
challenges related to the development, deployment, and maintenance of reusable
software.
%\note{Ketan: What are the levels in scientific process where software
%is required. It might be interesting to outline these stages and types of
%software required in each stage.}
In addition, it is essential that scientists,
researchers, and students are able to learn and adopt software-related skills
and methodologies. Established researchers are already acquiring some of these
skills, and in particular a specialized class of software developers is
emerging in academic environments as an integral part of
successful research teams. The WSSSPE workshop  provided a forum
for discussion of the challenges around scientific software, including both positions and experiences. The
workshop short papers and debates have been archived to provide a basis for continued
discussion, and have fed into the collaborative writing of this
document. Additional papers based on extended versions of workshop
submissions have been developed as part of this special journal edition. The level of interest in the workshop which involved 
an estimated 90 to 150 participants during the day has led to further activities:
Two additional WSSSPE workshops are planned  (at the 2014 SciPy and SC14 conferences), 
funds to support the workshops have been obtained from 
the National Science Foundation and the Gordon and Betty Moore
Foundation, and the
original workshop website has been turned into a community website for
further discussion and progress. Additionally, a minisymposium  to
further explore some of the key issues raised in WSSSPE is co-organized by a
WSSSPE1 participant at the 2014 Society for Industrial and Applied Mathematics
(SIAM) Annual Meeting on ``Reliable
Computational Science'' (SIAM AN14, \url{http://meetings.siam.org}).

This paper summarizes the  workshop papers
and discussion, and identifies issues that future WSSSPE events will address.
Before the WSSSPE1 workshop took place, the organizers self-published
a paper~\cite{WSSSPE1-pre-report} to document the process of
organizing and advertising the workshop, collecting and reviewing the
papers, and putting together the agenda. Section~\ref{sec:process} is
based on that publication. The remainder of this paper is based on the
workshop itself, as documented in a set of collaborative notes taken
during the workshop~\cite{WSSSPE1-google-notes}, including discussion
about the keynote presentations~(\S\ref{sec:keynotes}), the three
panels~(\S\ref{sec:devel}-\ref{sec:community}), 
cross-cutting issues~(\S\ref{sec:cross-cutting}),
use cases~(\S\ref{sec:use-cases}), and conclusions~(\S\ref{sec:conclusions}).  
Additionally, the accepted workshop papers are listed in
Appendix~\ref{sec:papers}, and a partial list of workshop attendees is
in Appendix~\ref{sec:attendees}.

@@

Another problem Bourne experienced was staff retention with respect to
private organizations which reward those combining research and
software expertise (the ``Google Bus''). However, he is a strong fan
of software sustainability through public-private partnerships. He
noted that making a successful business from scientific software alone
is very rare: founders overvalue while customers undervalue. He noted
that to last, an open source project needs a minimal funding
requirement even with a vibrant community --- goodwill only goes so
far if one is being paid to do something else.  He talked about grant
schemes of relevance in the U.S., particularly with regard to
technology transfer~\cite{sbir-web, fased-web}.

Bourne also had problems with selling research software: the university technology
transfer office wanted huge and unrealistic intellectual property
reach through, whereby they would get a share of profits from drugs
developed by pharmaceutical companies who use the software. 
He advocated for  a one-click approach for customers to
purchase university-written software.

@@

Bourne concluded by discussing his 2011 paper ``Ten Simple Rules for Getting Ahead as a Computational Biologist in Academia''~\cite{bourne_ten} and
argued that computational scientists ``have a responsibility to
convince their institutions, reviewers, and communities that software is
scholarship, frequently
more valuable than a research article''.

@@

The question of what ``sustainable'' means in the context of software was
raised (see \S\ref{sec:defining-sustainability}.) The resources at OSS Watch~\cite{osswatch}   discuss what
to look for when choosing software for procurement, or reuse in further software
development. Regardless of the license and development model that will be used,
the future of the project must be considered. Even if a particular piece of
software satisfies today's requirements, will it continue to do so in the
future? Important questions include whether the supplier will still be around in five
years' time, will it still care for its customers' needs, will it be responsive to
bug reports and feature requests? Should you tie your investment to a single
supplier using a proprietary product, or ensure the project uses an OSI-approved
license, and can outlive any single entity if the software is still important.

@@

The papers submitted to this panel included many general
recommendations for processes, practices, tools, etc. One of the papers
suggested that a ``Software Sustainability Institute'' should be vested with
the authority to develop standardized quality processes, a central common
repository, central resources for services and consulting, a think tank of
sorts, and a software orphanage center. The idea of one common repository received
some resistance, with so many compelling alternatives available, e.g. Bitbucket or
GitHub. The centralized communication or point of contact was seen as
reasonable, with the statement that ``vested with authority'' is perhaps too strong,
but ``providing tools if needed'' might be more appropriate.

@@

There was debate that while some of the basics of computing may be 
similar tools commonly used for computing have changed quite
significantly. Reference was made to Perl, which was commonly used, giving way
to whole new languages, such as Python, for gluing things together and how this
induces many students into entirely rewriting the scaffolding, leaving the old
to rot and the experiments to become non-reproducible as the tools change.
 There was
discussion of this tendency along with the enormous differences in the speed and
ease of sharing---having to ship tapes around (GCC, Emacs, etc) as opposed to
the immediate sharing of the latest development online, using revision control
systems like CVS, Subversion, Git, Mercurial, Bazaar, etc.
